### 2013-08-19 Docs/Thesis/background.tex
Final thesis revisions

The goal of FRP is to provide compositional and high-level abstractions for
creating reactive programs. The key abstractions of FRP are behaviors or
signals\footnote{Behaviors are generally referred to as {\em signals} in
signal-function literature. This is unfortunate, since a {\em signal function}
may operate on events, behaviors, or some combination of the two.}, which are
time-dependent values defined at every point in continuous time, and events,
which are values defined at countably many points in time. An FRP system will
provide combinators to manipulate events and behaviors and to react to events by
replacing a portion of the running program in response to an event. Behaviors
and events, or some abstraction based on them, will be first class, in keeping
with the spirit of functional programming. Programs implemented in FRP languages
should of course be efficiently executable, but this has proven to be the main
challenge in implementing FRP.

The two general approaches to FRP are ``classic'' FRP, where behaviors and
events are first-class and reactive objects, and ``signal-function'' FRP, where
transformers of signals and events are first-class and reactive objects.

@@

The first implementation of FRP outside the Haskell language was Frapp\'{e}, 
implemented in the Java Beans framework. Frapp\'{e} built on the notion
of events\footnote{These are semantic objects to which callbacks may be added,
and not events in the FRP sense, though they are related.} and ``bound
properties'' in the Beans framework, providing abstract interfaces for FRP
events and behaviors, and combinators as concrete classes implementing these
interfaces. The evaluation of Frapp\'{e} used Bean components as sources and
sinks, and the implementation of Bean events and bound properties to propagate
changes to the network~\cite{Courtney2001-2}.

@@

The FrTime\footnote{FrTime is available in the current version of Dr Racket (5.2.1).}~\cite{Cooper2006}
language extends the Scheme evaluator with a mutable dependency graph, which is
constructed by program evaluation. This graph is then updated by signal changes.
FrTime does not provide a distinct concept of events, and chooses branches of
the dependency graph by conditional evaluation of signals, rather than by the
substitution of signals used by FRP systems.

@@

Signal functions are time-dependent and reactive transformers of events and
signals. Signals and events cannot be directly manipulated by the programmer.
This approach has two motivations: it increases modularity since both the input
and output of signal functions may be transformed (as opposed to signals or
events which may only be transformed in their output) and it avoids a large
class of time and space leaks which have emerged when implementing FRP with
first-class signals and events.

@@

A recent PhD thesis~\cite{Sculthorpe2011} introduced N-Ary FRP, a technique for
typing Arrowized FRP systems using dependent types. The bulk of the work
consisted in using the dependent type system to prove the correctness of the FRP
system discussed. This work introduced signal vectors, a typing construct
which permits the distinction of signal and event types at the level of the FRP
system, instead of making events merely a special type of signal.

>>>>

The goal of FRP is to provide compositional and high-level abstractions for
creating reactive programs. The key abstractions of FRP are behaviors
\footnote{Behaviors are generally referred to as {\em signals} in
signal-function literature. This is unfortunate, since a {\em signal function}
may operate on events, behaviors, or some combination of the two.}, which are
time-dependent values defined at every point in continuous time, and events,
which are values defined at countably many points in time. An FRP system will
provide combinators to manipulate events and behaviors and to react to events by
replacing a portion of the running program in response to an event. Behaviors
and events, or some abstraction based on them, will be first class, in keeping
with the spirit of functional programming. Programs implemented in FRP languages
should of course be efficiently executable, but this has proven to be the main
challenge in implementing FRP.

The two general approaches to FRP are ``classic'' FRP, where behaviors and
events are first-class and reactive objects, and ``signal-function'' FRP, where
transformers of behaviors and events are first-class and reactive objects.

@@

The first implementation of FRP outside the Haskell language was Frapp\'{e}~\cite{Courtney2001-2},
implemented in the Java Beans framework. Frapp\'{e} built on the notion
of events\footnote{These are semantic objects to which callbacks may be added,
and not events in the FRP sense, though they are related.} and ``bound
properties'' in the Beans framework, providing abstract interfaces for FRP
events and behaviors, and combinators as concrete classes implementing these
interfaces. The evaluation of Frapp\'{e} used Bean components as sources and
sinks, and the implementation of Bean events and bound properties to propagate
changes to the network.

@@

The FrTime\footnote{FrTime is available in all recent versions of Dr. Racket}~cite{Cooper2006}
language extends the Scheme evaluator with a mutable dependency graph, which is
constructed by program evaluation. This graph is then updated by signal changes.
FrTime does not provide a distinct concept of events, and chooses branches of
the dependency graph by conditional evaluation of behaviors, rather than by the
substitution of behaviors used by FRP systems.

@@

Signal functions are time-dependent and reactive transformers of events and
behaviors. Behaviors and events cannot be directly manipulated by the programmer.
This approach has two motivations: it increases modularity since both the input
and output of signal functions may be transformed (as opposed to behaviors or
events which may only be transformed in their output) and it avoids a large
class of time and space leaks which have emerged when implementing FRP with
first-class behaviors and events.

@@

A recent PhD thesis~\cite{Sculthorpe2011} introduced N-Ary FRP, a technique for
typing Arrowized FRP systems using dependent types. The bulk of the work
consisted in using the dependent type system to prove the correctness of the FRP
system discussed. This work introduced signal vectors, a typing construct
which permits the distinction of behavior and event types at the level of the FRP
system, instead of making events merely a special type of behavior.

### 2013-08-19 Docs/Thesis/comparison.tex
Final thesis revisions

The example application from Chapter~\ref{chapter:Example_Application}, an
OpenFlow controller, was also implemented in Yampa. A slightly-modified
TimeFlies implementation, which used a polling loop rather than the GHC event
system to obtain events, but still "pushed" events to TimeFlies once they were
acquired, was also implemented.

These threee controllers were benchmarked using the {\tt cbench}
program~\cite{cbench}. The amount to time to pause between time and signal
updates (the inverse of the sampling rate for signals) was passed as a parameter
to each program. The {\tt cbench} benchmark operates by sending a simulated
packet-in event, waiting for a rule response, and then repeating. The output of
the benchmark is the number of times this occurs in a second. This is thus a
test of event latency in a reactive system.

>>>>

The example application from Chapter~\ref{chapter:Example_Application}, an
OpenFlow controller, was also implemented in Yampa. A slightly-modified
TimeFlies implementation, which used a polling loop rather than the GHC event
system to obtain events, but still ``pushed'' events to TimeFlies once they were
acquired, was also implemented.

These three controllers were benchmarked using the {\tt cbench}
program~\cite{cbench}. The amount to time to pause between time and signal
updates (the inverse of the sampling rate for signals) was passed as a parameter
to each program. The {\tt cbench} benchmark operates by sending a simulated
packet-in event, waiting for a rule response, and then repeating. The output of
the benchmark is the number of times this occurs in a second. This is thus a
test of event latency in a reactive system.

### 2013-08-19 Docs/Thesis/conclusions-and-further-work.tex
Final thesis revisions

Many classes of reactive application would benefit from a ``dynamic collections'' combinator
similar to {\tt pSwitch} in Yampa. Such a combinator allows a collection of signal functions
to be evaluated as one signal function, with addition or removal of signal functions instead
of the total replacement given by the {\tt switch} combinator. This is useful, for instance,
when simulating objects for games or computer models, as the behavior of each object can be
modeled as a signal function, and these signal functions can be added to and removed from the
collection. An attempt was made to add such a function to TimeFlies, but it is not clear what 
the type of such a combinator would be in a system employing signal vectors, or how the
"routing" of inputs to individual signal functions in the collection would be specified.

>>>>

Many classes of reactive application would benefit from a ``dynamic collections'' combinator
similar to {\tt pSwitch} in Yampa. Such a combinator allows a collection of signal functions
to be evaluated as one signal function, with addition or removal of signal functions instead
of the total replacement given by the {\tt switch} combinator. This is useful, for instance,
when simulating objects for games or computer models, as the behavior of each object can be
modeled as a signal function, and these signal functions can be added to and removed from the
collection. An attempt was made to add such a function to TimeFlies, but it is not clear what 
the type of such a combinator would be in a system employing signal vectors, or how the
``routing'' of inputs to individual signal functions in the collection would be specified.

### 2013-08-19 Docs/Thesis/example.tex
Final thesis revisions

The first component for our learning switch is the table which maps addresses
to ports. This is a stateful data structure which will be updated by input
events and possibly produce output events. We employ the {\tt accumulateList}
combinator to produce the signal function:

@@

The other task is to periodically eliminate all rules which are older than a
specified threshold. For this purpose, we construct another signal function
which produces {\tt TableAccumulator} events. This signal function runs a timer
(the {\tt every} combinator) and produces events carrying closures which scan
the table for old rules, delete them, and produces messages to inform the
switches of their deletion. This signal function is shown in Figure~\ref{figure:clean_sf},
again eliding the implementation of the closures.

@@

\begin{figure}
\begin{code}
-- | Entry point
main :: IO ()
main = do (argS:_) <- getArgs
          let sampleTime = read argS
          time <- getMonoTimeAsDouble
          reactiveRef <- newIORef $ initSFEval 
                                      (eventHandler id)
                                      (combineSamples sampleEvt sampleEvt)
                                      time
                                      learningSwitch
          ofEventManager <- 
            openFlowEventManager
              Nothing
              port 
              -- Switch addition inputs
              (\handle -> do rState <- readIORef reactiveRef
                             ((), rState') <- runSFEvalT 
                                                (push $ svLeft $ svOcc handle)
                                                rState
                             writeIORef reactiveRef rState')
              -- Switch message inputs:
              (\((tid, msg), handle) -> 
                 do rState <- readIORef reactiveRef
                              ((), rState') <- runSFEvalT 
                                                 (push $ svRight $
                                                  svOcc (handle, tid, msg))
                                                 rState
                              writeIORef reactiveRef rState')
          let evtMgr = getEventManager ofEventManager
              -- Sampling:
              sample = void $ registerTimeout evtMgr sampleTime $ 
                         do rState <- readIORef reactiveRef
                            time <- getMonoTimeAsDouble
                            ((), rState') <- runSFEvalT (step time) rState
                            writeIORef reactiveRef rState'
                            sample
          sample 
          loop $ evtMgr      
\end{code}
\hrule
\caption{Using the evaluation interface for IO.}
\label{figure:example_io}
\end{figure}

>>>>

The first component for our learning switch is the table which maps addresses
to ports. This is a stateful data structure which will be updated by input
events and possibly produce output events. We employ the {\tt accumulateList}
combinator to produce the signal function shown in Figure~\ref{figure:switch_table_sf}.

@@

The other task is to periodically eliminate all rules which are older than a
specified threshold. For this purpose, we construct another signal function
which produces {\tt TableAccumulator} events. This signal function runs a timer
(the {\tt every} combinator, which is implemented using {\tt switch} and {\tt after}
and produces an event repeatedly on a given interval) and produces events carrying closures which scan
the table for old rules, delete them, and produces messages to inform the
switches of their deletion. This signal function is shown in Figure~\ref{figure:clean_sf},
again eliding the implementation of the closures.

@@

\begin{figure}
\begin{code}
-- | Entry point
main :: IO ()
main = do (argS:_) <- getArgs
          let sampleTime = read argS
          time <- getMonoTimeAsDouble
          reactiveRef <- newIORef $ initSFEval 
                                      (eventHandler id)
                                      (combineSamples sampleEvt sampleEvt)
                                      time
                                      learningSwitch
          ofEventManager <- 
            openFlowEventManager
              Nothing
              port 
              -- Switch addition inputs
              (\handle -> do rState <- readIORef reactiveRef
                             ((), rState') <- runSFEvalT 
                                                (push $ svLeft $ svOcc handle)
                                                rState
                             writeIORef reactiveRef rState')
              -- Switch message inputs:
              (\((tid, msg), handle) -> 
                 do rState <- readIORef reactiveRef
                    ((), rState') <- runSFEvalT 
                                       (push $ svRight $
                                        svOcc (handle, tid, msg))
                                       rState
                    writeIORef reactiveRef rState')
          let evtMgr = getEventManager ofEventManager
              -- Sampling:
              sample = void $ registerTimeout evtMgr sampleTime $ 
                         do rState <- readIORef reactiveRef
                            time <- getMonoTimeAsDouble
                            ((), rState') <- runSFEvalT (step time) rState
                            writeIORef reactiveRef rState'
                            sample
          sample 
          loop evtMgr      
\end{code}
\hrule
\caption{Using the evaluation interface for IO.}
\label{figure:example_io}
\end{figure}

### 2013-08-19 Docs/Thesis/implementation-signalfunctions.tex
Final thesis revisions

The types are therefore expressed in the system exactly as they were described
in Chapter~\ref{chapter:System_Design_and_Interface}. (To refresh, see
Figure~\ref{figure:signal_vector_types}.) The striking observation about these
types is that they have {\em no data constructors}. There are no values which
take these types.

@@

The {\tt switch} combinator stores the input sample provided during initialization,
and updates it with the input signal updates. When the wrapped signal function
produces an occurrence carrying a new signal function, that signal function is
initialized with the stored input sample. It is then "wrapped" by another
function which closes over its output sample, and outputs the sample as a signal
update as the next time step. After this, it acts as the new signal function.
This wrapping has some performance implications, which are discussed

>>>>

The types are therefore expressed in the system exactly as they were described
in Chapter~\ref{chapter:System_Design_and_Interface}. (To refresh, see
Figure~\ref{figure:signal_vector_types}.) The striking observation about these
types is that they have {\em no data constructors}. There are no values which
have these types.

@@

The {\tt switch} combinator stores the input sample provided during initialization,
and updates it with the input signal updates. When the wrapped signal function
produces an occurrence carrying a new signal function, that signal function is
initialized with the stored input sample. It is then ``wrapped'' by another
function which closes over its output sample, and outputs the sample as a signal
update as the next time step. After this, it acts as the new signal function.
This wrapping has some performance implications, which are discussed in Chapter~\ref{chapter:Evaluation_and_Comparisons}.

### 2013-08-19 Docs/Thesis/system_design_and_interface.tex
Final thesis revisions

It is fine for a system to be composable with regards to itself, but an FRP
system must interact with the outside world. Since we cannot anticipate every
possible form of input and output that the system will be asked to interact
with, we must interface with Haskell's IO system. In particular, most libraries
for user interaction (e.g. GUI and graphics libraries such as GTK+ and GLUT) and
most libraries for time-dependent IO (e.g. audio and video systems) make use of
the event loop abstraction. In this abstraction, event handlers are registered
with the system, and then a command is issued to run a loop which detects events
and runs the handlers, and uses the results of the handlers to render the
appropriate output.

@@

TimeFlies provides a monad transformer for evaluating signal functions. Thi
monad transformer may be used in conjunction with the {\tt IO} monad (or any
other monad) to describe how input is to be obtained for the signal function
being evaluated, and how outputs are to be handled.

@@

\begin{figure}
\begin{code}
-- A signal function's evaluations state
data SFEvalState svIn svOut m
-- The evaluation monad
data SFEvalT svIn svOut m a

@@

With the exception of {\tt identity}, all of the basic signal functions have
empty inputs. This allows these signal functions to be used to insert values
into the network which are known when the signal function is created, without
having to route those values from an input.

@@

-- Update the value of an input signal sample
-- (not immediately observed)
update :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()

>>>>

It is fine for a system to be composable with regards to itself, but an FRP
system must interact with the outside world. Since we cannot anticipate every
possible form of input and output that the system will be asked to interact
with, we must interface with Haskell's IO system. In particular, most libraries
for user interaction (e.g. GUI and graphics libraries such as GTK+ and GLUT) and
most libraries for time-dependent IO (e.g. audio and video systems) make use of
the event-loop abstraction. In this abstraction, event handlers are registered
with the system, and then a command is issued to run a loop which detects events
and runs the handlers, and uses the results of the handlers to render the
appropriate output.

@@

TimeFlies provides a monad transformer for evaluating signal functions. This
monad transformer may be used in conjunction with the {\tt IO} monad (or any
other monad) to describe how input is to be obtained for the signal function
being evaluated, and how outputs are to be handled.

@@

\begin{figure}
\begin{code}
-- A signal function's evaluation state
data SFEvalState svIn svOut m
-- The evaluation monad
data SFEvalT svIn svOut m a

@@

With the exception of {\tt identity}, all of the basic signal functions have
empty inputs. This allows these signal functions to be used to insert values
into the network which are known when the signal function is created, without
having to route those values from an external input.

@@

-- Update the value of an input signal sample
-- (not immediately observed)
update :: (Monad m) => SVSignalUpdate svIn -> SFEvalT svIn svOut m ()
